Containers and R---the Rockerverse and beyond

### 2019-12-13 manuscript.Rmd

The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.

To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
For any R projects that use the Tidyverse suite [citation in review. Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.

@@

One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
The former models provide modularity, while the latter approach is most similar to a simple shared server.
In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
The Rocker project provides containers pre-installed with the RStudio IDE.

In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.

In some cases, containers are not completely portable between hardware environments.
This is the case for software using graphical processing units (GPUs) which are increasingly popular for compute-intensive machine learning tasks.
Containers running GPU software  require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugn for docker and set of base images with appropriate drivers and libraries, and the Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages.

R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database. 
But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by the IT. 
Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a docker image that contains everything needed to connect, from R, to a database. 
Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a docker instance of a specific database, and how to use `r-db` to connect to it. 
Each package contained inside this image also has a series examples, so that the user can get started with the database right away.

- [RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user

>>>>

The fact that Docker images are portable and well defined make them useful when more than one person needs access to the same computing environment.
This is even more useful when some of the users do not have the expertise to create such an environment themselves, and when these environments can be run in public or shared infrastructure.

[`holepunch`](https://github.com/karthik/holepunch) is an R package that was designed to make sharing work environments accessible to novice R users based on _Binder_.
The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
Binder's core internal tool is [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html).
It deterministically builds a Docker image by parsing the contents of a repository, e.g. project dependency configurations or simple configuration files.
In the most powerful case, `repo2docker` builds a given Dockerfile.
While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
For any R projects that use the Tidyverse suite [citation in review, will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
`holepunch` removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
`holepunch` as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
The Dockerfile included with the project can also be used to launch a RStudio server locally independent of Binder which is especially useful when more or special computational resources are required.

@@

In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
The former models provide modularity, while the latter approach is most similar to a simple shared server.
Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
The Rocker project provides containers pre-installed with the RStudio IDE.
In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible [REF].
GPUs are increasingly popular for compute-intensive machine learning tasks [REF].
Though in this case, containers are not completely portable between hardware environments.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.

**Teaching** is a further example where sharing a prepared computing environment can greatly improve the process, especially for courses that require access to a relatively complex setup of software tools, e.g., as in the case of database systems.
R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database. 
But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by IT services.
Giving access to a sandbox for the most common database environments is the idea behind **[`r-db`](https://github.com/ColinFay/r-db)** [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a Docker instance of a specific database, and how to use `r-db` to connect to it.
Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.

**[RCloud](https://github.com/att/rcloud/tree/master/docker)** uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
